Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-03-01 paper/main.tex
final edits to methods and ending sections of paper

We enrolled a total of 50 Dartmouth undergraduate students in our study.
Participants received course credit for enrolling. We asked each participant to
complete a demographic survey that included questions about their age, gender,
native spoken language, ethnicity, race, hearing, color vision, sleep, coffee
consumption, level of alertness, and several aspects of their educational
background and prior coursework.

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing Lecture~1, the second between
Lectures~1 and 2, and the third after viewing Lecture~2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about Lecture~1, 5 questions about
Lecture~2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model (Latent
Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set of documents, to
discover a set of $k$ ``topics'' or ``themes.'' Formally, each topic is defined
as a set of weights over each word in the model's vocabulary (i.e., the union
of all unique words, across all documents, excluding ``stop words.'').
Conceptually, each topic is intended to give larger weights to words that are
semantically related or tend to co-occur in the same documents. After
fitting a topic model, each document in the training set, or any \textit{new}
document that contains at least some of the words in the model's vocabulary,
may be represented as a $k$-dimensional vector describing how much the document
(most probably) reflects each topic.
To select an appropriate $k$ for our model, we identified the minimum number of
topics that yielded at least one ``unused'' topic (i.e., in which all words in the vocabulary
were assigned zero weight) after training. This indicated that the number of topics was
sufficient to capture the set of latent themes present in the two lectures (from which we
constructed our document corpus, as described below). We found this value to be $k = 15$
topics. The distribution of weights over words in the vocabulary for each discovered topic is
shown in Supplementary Figure~\topicWordWeights, and each topic's top-weighted
words may be found in Supplementary Table~\topics.

As illustrated in Figure~\ref{fig:sliding-windows}A, we start by building up a
corpus of documents using overlapping sliding windows that span each video's
transcript. Khan Academy provides professionally created, manual transcriptions
of all videos for closed captioning. However, such transcripts would not be
readily available in all contexts to which our framework could potentially be
applied. Khan Academy videos are hosted on the YouTube platform, which
additionally provides automated captions. We opted to use these automated
transcripts (which, in prior work, we have found are of sufficiently near-human
quality yield reliable data in behavioral studies; \citealp{ZimaEtal18}) when
developing our framework in order to make it more directly extensible and
adaptable by others in the future.

We fetched these automated transcripts using the
\texttt{youtube-transcript-api} Python package~\citep{Depo19}. The transcripts
consisted of one timestamped line of text for every few seconds (mean: 2.34~s;
standard deviation: 0.83~s) of spoken content in the video (i.e., corresponding
to each individual caption that would appear on-screen if viewing the lecture
via YouTube, and when those lines would appear). We defined a sliding window
length of (up to) $w = 30$ transcript lines, and assigned each window a
timestamp corresponding to the midpoint between its first and last lines'
timestamps. These sliding windows ramped up and down in length at the very
beginning and end of the transcript, respectively. In other words, the first
sliding window covered only the first line from the transcript; the second
sliding window covered the first two lines; and so on. This insured that each
line of the transcript appeared in the same number ($w$) of sliding windows.
After performing various standard text preprocessing (e.g., normalizing case,
lemmatizing, removing punctuation and stop-words), we treated the text from
each sliding window as a single ``document,'' and combined these documents
across the two videos' windows to create a single training corpus for the topic
model. The top words from each of the 15 discovered topics may be found in
Supplementary Table~\topics.

@@

We transformed each sliding window's text into a topic vector, and then used
linear interpolation (independently for each topic dimension) to resample the
resulting timeseries to one vector per second. We also used the fitted model to
obtain topic vectors for each question in our pool (Supplementary Tab.~\questions). Taken
together, we obtained a \textit{trajectory} for each video, describing its path
through topic space, and a single coordinate for each question
(Fig.~\ref{fig:sliding-windows}C). Embedding both videos and all of the
questions using a common model enables us to compare the content from different
moments of videos, compare the content across videos, and estimate potential
associations between specific questions and specific moments of video.

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions or even
appearing in the lectures. To visualize these estimates
(Fig.~\ref{fig:knowledge-maps}, Supplementary Figs.~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection (UMAP; \citealp{McInEtal18a}) to construct a 2D projection of the
text embedding space. Sampling the original 100-dimensional space at high
resolution to obtain an adequate set of topic vectors spanning the embedding
space would be computationally intractable. However, sampling a 2D grid is
trivial.

At a high level, the UMAP algorithm obtains low-dimensional embeddings by
minimizing the cross-entropy between the pairwise (clustered) distances between
the observations in their original (e.g., 100-dimensional) space and the
pairwise (clustered) distances in the low-dimensional embedding space (in our
approach, the embedding space is 2D). In our implementation, pairwise distances
in the original high-dimensional space were defined as 1 minus the correlation
between the pair of coordinates, and pairwise distances in the low-dimensional
embedding space were defiend as the Euclidean distance between the pair of
coordinates.

In our application, all of the coordinates we embedded were topic vectors,
whose elements are always non-negative. Although UMAP is an invertible
transformation at the embedding locations of the original data, other locations
in the embedding space will not necessarily follow the same implicit ``rules''
as the original high-dimensional data. For example, inverting an arbitrary
coordinate in the embedding space might result in negative-valued vectors,
which are incompatable with the topic modeling framework. To protect against
this issue, we log-transformed the topic vectors prior to embedding them in the
2D space. When we inverted the embedded vectors (e.g., to estimate topic
vectors or word clouds, as in Fig.~\ref{fig:knowledge-maps}C), we passed the
inverted (log-transformed) values through the exponential function to obtain a
vector of non-negative values.

>>>>

We enrolled a total of 50 Dartmouth undergraduate students in our study.
Participants received optional course credit for enrolling. We asked each participant to
complete a demographic survey that included questions about their age, gender,
native spoken language, ethnicity, race, hearing, color vision, sleep, coffee
consumption, level of alertness, and several aspects of their educational
background and prior coursework.

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing Lecture~1, the second between
Lectures~1 and 2, and the third after viewing Lecture~2
(see Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain exactly 5 questions about Lecture~1, 5 questions about
Lecture~2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used this experiment to
develop and test our computational framework for estimating knowledge and
learning.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model (Latent
Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set of documents, to
discover a set of (up to) $k$ ``topics'' or ``themes.'' Formally, each topic is defined
as a distribution of weights over each word in the model's vocabulary (i.e., the union
of all unique words, across all documents, excluding ``stop words.'').
Conceptually, each topic is intended to give larger weights to words that are
semantically related or tend to co-occur in the same documents. After
fitting a topic model, each document in the training set, or any \textit{new}
document that contains at least some of the words in the model's vocabulary,
may be represented as a $k$-dimensional vector describing how much the document
(most probably) reflects each topic.
To select an appropriate $k$ for our model, we identified the minimum number of
topics that yielded at least one ``unused'' topic (i.e., in which all words in the vocabulary
were assigned uniform weights) after training. This indicated that the number of topics was
sufficient to capture the set of latent themes present in the two lectures (from which we
constructed our document corpus, as described below). We found this value to be $k = 15$
topics. The distribution of weights over words in the vocabulary for each discovered topic is
shown in Supplementary Figure~\topicWordWeights, and each topic's top-weighted
words may be found in Supplementary Table~\topics.

As illustrated in Figure~\ref{fig:sliding-windows}A, we start by building up a
corpus of documents using overlapping sliding windows that span each video's
transcript. Khan Academy provides professionally created, manual transcriptions
of all videos for closed captioning. However, such transcripts would not be
readily available in all contexts to which our framework could potentially be
applied. Khan Academy videos are hosted on the YouTube platform, which
additionally provides automated captions. We opted to use these automated
transcripts (which, in prior work, we have found to be of sufficiently near-human
quality to yield reliable data in behavioral studies;~\citealp{ZimaEtal18}) when
developing our framework in order to make it more directly extensible and
adaptable by others in the future.

We fetched these automated transcripts using the
\texttt{youtube-transcript-api} Python package~\citep{Depo18}. The transcripts
consisted of one timestamped line of text for every few seconds (mean: 2.34~s;
standard deviation: 0.83~s) of spoken content in the video (i.e., corresponding
to each individual caption that would appear on-screen if viewing the lecture
via YouTube, and when those lines would appear). We defined a sliding window
length of (up to) $w = 30$ transcript lines, and assigned each window a
timestamp corresponding to the midpoint between the timestamps for its first and last lines.
These sliding windows ramped up and down in length at the
beginning and end of each transcript, respectively. In other words, each transcript's first
sliding window covered only its first line, the second
sliding window covered the first two lines, and so on. This ensured that each
line from the transcripts appeared in the same number ($w$) of sliding windows.
%and was equally represented in the topic model's training corpus.
After performing various standard text preprocessing (e.g., normalizing case,
lemmatizing, removing punctuation and stop-words), we treated the text from
each sliding window as a single ``document,'' and combined these documents
across the two videos' windows to create a single training corpus for the topic model.
%To select an appropriate number of topics for the model, we identified the
%minimum $k$ that yielded at least one ``unused'' topic (i.e., in which all words
%in the vocabulary were assigned zero weight) after training. This indicated that the
%number of topics was sufficient to capture the set of latent themes present in the two
%lectures. We found this value to be $k = 15$ topics.
%Supplementary Figure~\topicWordWeights~displays the distribution of weights over words
%in the vocabulary for each discovered topic, and each topic's top-weighted words may be found
%in Supplementary Table~\topics.

@@

We transformed each sliding window's text into a topic vector, and then used
linear interpolation (independently for each topic dimension) to resample the
resulting timeseries to one vector per second. We also used the fitted model to
obtain topic vectors for each question in our pool (see Supplementary Tab.~\questions). Taken
together, we obtained a \textit{trajectory} for each video, describing its path
through topic space, and a single coordinate for each question
(Fig.~\ref{fig:sliding-windows}C). Embedding both videos and all of the
questions using a common model enables us to compare the content from different
moments of videos, compare the content across videos, and estimate potential
associations between specific questions and specific moments of video.

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions, or even
appearing in the lectures. To visualize these estimates
(Fig.~\ref{fig:knowledge-maps}, Supplementary Figs.~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection (UMAP; \citealp{McInEtal18a, McInEtal18b}) to construct a 2D projection of the
text embedding space. Sampling the original 100-dimensional space at high
resolution to obtain an adequate set of topic vectors spanning the embedding
space would be computationally intractable. However, sampling a 2D grid is
trivial.

At a high level, the UMAP algorithm obtains low-dimensional embeddings by
minimizing the cross-entropy between the pairwise (clustered) distances between
the observations in their original (e.g., 100-dimensional) space and the
pairwise (clustered) distances in the low-dimensional embedding space (in our
approach, the embedding space is 2D). In our implementation, pairwise distances
in the original high-dimensional space were defined as 1 minus the correlation
between each pair of coordinates, and pairwise distances in the low-dimensional
embedding space were defined as the Euclidean distance between each pair of
coordinates.

In our application, all of the coordinates we embedded were topic vectors,
whose elements are always non-negative and sum to one. Although UMAP is an invertible
transformation at the embedding locations of the original data, other locations
in the embedding space will not necessarily follow the same implicit ``rules''
as the original high-dimensional data. For example, inverting an arbitrary
coordinate in the embedding space might result in negative-valued vectors,
which are incompatible with the topic modeling framework. To protect against
this issue, we log-transformed the topic vectors prior to embedding them in the
2D space. When we inverted the embedded vectors (e.g., to estimate topic
vectors or word clouds, as in Fig.~\ref{fig:knowledge-maps}C), we passed the
inverted (log-transformed) values through the exponential function to obtain a
vector of non-negative values, and normalized them to sum to one.
